A New Approach for Parallel Functional Arrays

### 2016-05-05 Paper/paper.tex
Add author
Syncing github with other repo

@@

\authorinfo{Ananya Kumar}
           {Carnegie Mellon University}
           {ananyak@andrew.cmu.edu}
\authorinfo{Guy Blelloch}
           {Carnegie Mellon University}
           {blelloch@cs.cmu.edu}

\maketitle

@@

\section{Introduction}

Arrays are very important in functional programming languages because they allow work-efficient implementations of algorithms like depth-first search. Accessing old versions of arrays can be useful for efficient checkpointing, logging, and event handling.

\subsection{Previous Approaches}

There has been a lot of great work on functional arrays, but previous approaches have limitations.

In many functional programming languages, like Haskell, monads ~\cite{wadler-monads} can be used to enforce single threaded use of arrays. However, enforcing single threaded use obstructs persistence - it prevents us from accessing or modifying old versions of an array.

Another approach is to use reference counting ~\cite{hudak}. If the number of references to an array is one, the array can be updated in place, otherwise the array is copied and then updated. This approach makes it difficult for programmers to statically analyze the time complexity of array operations. Furthermore, even if there are multiple reference to an array the array might not need to be copied, so the unnecessary copying could make the implementation inefficient.

A third approach is to use linear types ~\cite{wadler-lintypes} to ensure that programmers can only access and mutate arrays in valid ways. Linear types have not been used practically for a few decades since its inception, and would require a complete re-think of a language's type system to be useful.

O' Neill's ~\cite{oneill} functional arrays guarantee $O(1)$ access and insertions when used like an imperative array. However, the arrays explicitly implement version trees which are inefficient in practice. Additionally, the reads and writes to the most recent version of the array might take $O(\log{n})$ work if the array is not used in a single threaded way.

Chuang's approach ~\cite{chuang} supports $O(1)$ accesses and insertions to the most recent version of arrays, however accessing old versions of the array takes $O(n)$ work which is often impractical.

None of the existing approaches support concurrent operations on arrays. O' Neill suggests having a lock for each element of the array. However, when many threads contend for the same element, this would serialize accesses and make accesses not $O(1)$. Additionally, per-element locks add significant memory overhead. Separation logic ~\cite{reynolds} could be used to parallelize divide and conquer algorithms on arrays however they would severely restrict the way functional arrays can be used.

\subsection{Our Approach}

Our arrays support 3 functions: tabulate, get, and set. The dynamics are functional, in particular set returns a new array and leaves the original array intact.

\begin{figure}[!ht]
\centering
\includegraphics[scale=0.45]{leaf_interior_intro}
\nocaptionrule \caption{Example usage of functional arrays}
\label{fig:leaf_interior_intro}
\end{figure}

After all the operations performed in figure \ref{fig:leaf_interior_intro}, A and C are \emph{interior} arrays and B and D are \emph{leaf} arrays. Calling get and set on leaf arrays costs $O(1)$ but calling get and set on interior arrays is more expensive. In the cost dynamics, we thread a store that keeps track of whether an array is an interior or leaf array.

We also give and prove fork join cost dynamics for our arrays. The dynamics for arrays used concurrently are non trivial because one thread might call get on a leaf array while another thread calls set. To deal with this, we compute the work of each fork separately, and then charge additional work at the join point if an array is used in both forks.

In our implementation of a functional array, we keep an array of values (multiple functional arrays might point to the same array). For each index of the array, we also keep a history of values that were previously at that index. To access an element in an interior array, we need to search through the history of values.

We give a wait-free concurrent implementation that uses careful synchronization. We prove linearizability of the implementation and use that to prove that our arrays work correctly when used concurrently. Finally, we give a provable implementation for our cost dynamics.

@@

We use a standard applicative-order language defined as follows:
$$e = x \; | \; c  \; | \; \lambda x.e \; | \; e_1 e_2 \; | \; e_1 || e_2 \; | \; \text{if } e_1 \text{ then } e_2 \text{ else } e_3$$

$L$ contains the usual arithmetic types, such as the natural numbers, and numerical operations such as sums and products. Lambda expressions and constants (including arrays) are terminal values in our language. Our dynamics are given by the judgement $e \Downarrow v$ which means that expression $e$ evaluates to $v$. Dynamics for function application and fork-join are given as examples.

$$\frac{e_1 \Downarrow v_1 \;\;\; e_2 \Downarrow v_2}{e_1 || e_2 \Downarrow (v_1, v_2)} \text{ (fork-join)}$$
$$\frac{f \Downarrow \lambda x . e \;\;\; y \Downarrow v \;\;\; [v/x]e \Downarrow v'}{f \; y \Downarrow v'} \text{ (function-app)}$$

The language also includes 3 functions for working with arrays: tabulate, get, and set. The dynamics for tabulate is given below:
$$\frac{f(1) \Downarrow v_1, ..., f(n) \Downarrow v_n}{\text{tabulate} \; f \; n \Downarrow [1 \mapsto v_1, ..., n \mapsto v_n]}$$

$get(A,i)$ returns the value at the $i$th index of $A$. $set(A,i,v)$ returns a new array where the value at index $i$ is $v$ and the value at all other indices is the same as in $A$. The evaluation of a simple program in our language is shown below:\\

$(\lambda A. \; get(A, 2) + get(A,3)) set(tabulate(\lambda i . \; i, 5), 2, 10)$

$(\lambda A. \; get(A, 2) + get(A,3)) set([1,2,3,4,5], 2, 10)$

$(\lambda A. \; get(A, 2) + get(A,3)) [1,10,3,4,5]$

$get([1,10,3,4,5], 2) + get([1,10,3,4,5],3) $

@@

The dynamics given in the previous section are functional. However, the costs of get and set depend on whether the array is a leaf array or an internal array. To deal with this, we thread a store through the cost dynamics to keep track of whether each array is a leaf or internal array.

@@

$$\frac{\delta; e_1 \Downarrow \delta'; \lambda x . e; w_1 ; s_1 \;\;\; \delta'; e_2 \Downarrow \delta'';v; w_2 ; s_2 \;\;\; \delta''; [v/x]e \Downarrow \delta'''; v'; w_3 ; s_3}{\delta; e_1 \; e_2 \Downarrow \delta'''; v'; 1+w_1+w_2+w_3; 1+s_1+s_2+s_3}$$

@@

Arrays are represented by $(l, V)$ where $l$ is a label that uniquely identifies the array, and $V$ contains the values in the array. $V[i]$ is the $i^{\text{th}}$ value in the array. If the set function has been called on an array, the array is an interior array (represented by $-$), otherwise the array is a leaf array (represented by $+$).

Array functions have different costs depending on whether the array is a leaf or interior array. Let $C(l)$ be the number of get operations applied on array $A = (l, V)$ that had $O(1)$ work. We thread a store $\delta$ through each operation. $\delta$ maps array labels $l$ to $(+/-, c)$, where $c = C(l)$. Let $f(n)$ be the work of calling get on an interior array and $g(n)$ be the work of calling set on an interior array.

The cost dynamics for the functions on a functional array of size $n$ are:
$$\frac{l = \text{new label} \;\;\; f(1) \Downarrow v_1; w_1; s_1; ... f(n) \Downarrow v_{n}; w_{n}; s_{n}}{\delta; \text{tabulate } f \; n \Downarrow \delta[l \mapsto (+, 0)]; (l, \bigcup_{i=1}^n [i \mapsto v_i]); 1+\sum w_i; 1+\max s_i}$$
$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (+, c)] \;\;\; l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l \mapsto (-, c), l' \mapsto (+, 0)]; (l', V[i \mapsto v]); 1; 1} \text{  (set-leaf)}$$
$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (-, c)] \;\;\;  l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l' \mapsto (+, 0)]; (l', V[i \mapsto v]); g(n); 1} \text{  (set-interior)}$$
$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (+, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta[l \mapsto (+, c+1)]; V[i]; 1; 1} \text{  (get-leaf)}$$
$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (-, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta; V[i]; f(n); 1} \text{  (get-interior)}$$

Set creates a new label and array and modifies the store to indicate that the newly created array is a leaf array, and the array set was called on has become an interior array. The work for set is higher for interior arrays. Get returns the value at the desired index, if the array is a leaf then the operation costs $O(1)$ but we increment the counter of $O(1)$ gets in the store.

@@

We assume that newly created labels in $\delta_L$ and $\delta_R$ do not conflict. Consider an array $A = (l, V)$ with $\delta[l] = (s,c)$, $\delta_L[l] = (s_L, c + c_L)$, $\delta_R[l] = (s_R, c + c_R)$. When multiplying two signs or multiplying a sign with an integer, consider + to be 1 and - to be 0.

Combine describes how to combine the values in the store on both sides of the fork-join. The array is a leaf array iff it is a leaf array on both sides of the fork-join. $O(1)$ gets on one side of the fork remain $O(1)$ iff there were no calls to set on the other side of the fork.
\begin{equation*}
  \begin{aligned}
   combine&((s, c), (s_L, c + c_L), (s_R, c + c_R))= \\
   &(s_L s_R, c + s_R c_L + s_L c_R)
  \end{aligned}
\end{equation*}
Extrawork gives the additional cost incurred if an array was modified on either side of a fork-join. If the array was an interior array before the fork-join, then all functions incurred their maximal cost and there is no additional cost. Otherwise, $O(1)$ gets on one side of the fork become $O(f(n))$ iff the other side of the fork called set. If both sides of the fork called set, then one of the sets came first and is $O(1)$, the subsequent set is $O(g(n))$.
\begin{equation*}
  \begin{aligned}
  extrawork&((s, c), (s_L, c + c_L), (s_R, c + c_R)) = \\
  &s \cdot ( (\neg s_R) c_L f(n)  + (\neg s_L) c_R f(n) + (\neg s_R) (\neg s_L) g(n) )
 \end{aligned}
\end{equation*}

\begin{equation*}
  \begin{aligned}
  \delta' = \; &(\delta_L / \delta) \; \cup \; (\delta_R / \delta) \; \cup \\
               &\bigcup_{l \in L(\delta_L) \cap L(\delta_R)} [l \mapsto combine(\delta[l], \delta_L[l], \delta_R[l])]
  \end{aligned}
\end{equation*}
$$w' = \sum_{l \in L(\delta_1) \cap L(\delta_2)} extrawork(\delta[l], \delta_L[l], \delta_R[l])$$
Then, the fork-join cost semantics are:
$$\delta; e_1 || e_2 \Downarrow \delta'; (v_1, v_2); 1 + w_1 + w_2 + w'; 1 + \max(s_1, s_2)$$

>>>>

\newcommand{\func}[1]{{\sc #1}}
\newcommand{\tabulate}{\func{tabulate}}
\newcommand{\get}{\func{get}}
\newcommand{\set}{\func{set}}
\newcommand{\tget}{\text{\get}}
\newcommand{\tset}{\text{\set}}
\newcommand{\ttab}{\text{\tabulate}}

@@

\authorinfo{Ananya Kumar}
           {Carnegie Mellon University}
           {ananyak@andrew.cmu.edu}
\authorinfo{Guy Blelloch}
           {Carnegie Mellon University}
           {blelloch@cs.cmu.edu}
\authorinfo{Robert Harper}
           {Carnegie Mellon University}
           {rwh@cs.cmu.edu}
\maketitle

@@

\input{intro}

@@

We use a standard applicative-order source language defined as follows:
$$e = x \; | \; c  \; | \; \lambda x.e \; 
| \; e_1 e_2 \; 
| \; (e_1, e_2) \; 
| \; (e_1 || e_2) \; 
| \; \text{if } e_1 \text{ then } e_2 \text{ else } e_3$$

The constants $c$ contains the usual arithmetic types, such as the
natural numbers and numerical functions.  The expression $(e_1||e_2)$
indicates that the two expressions can run in parallel, returning a
pair when both expressions are fully evaluated, and $(e_1,e_2)$
generates a pair sequentially.  The terminal (fully evaluated) values
are lambda expressions, as well as pairs and arrays of terminal
values, and are denoted by the judgement val.   Our value semantics are given by the judgement:
\[e \Downarrow v\]
indicating that the expression $e$ evaluates to $v$.  The rules for
function application and parallel pair are given as examples, and
other rules are standard.
$$\frac{e_1 \Downarrow v_1 \;\;\; e_2 \Downarrow v_2}{e_1 || e_2 \Downarrow (v_1, v_2)} \text{ (fork-join)}$$
$$\frac{f \Downarrow \lambda x . e \;\;\; y \Downarrow v \;\;\; [v/x]e \Downarrow v'}{f \; y \Downarrow v'} \text{ (function-app)}$$

The language also includes three functions for working with arrays: \tabulate{}, \get{}, and \set{}. The dynamics for \tabulate{} is given below:
$$\frac{f(1) \Downarrow v_1, ..., f(n) \Downarrow v_n}{\text{\tabulate{}} \; f \; n \Downarrow [1 \mapsto v_1, ..., n \mapsto v_n]}$$

\get$(A,i)$ returns the value at the $i$th index of $A$. \set$(A,i,v)$ returns a new array where the value at index $i$ is $v$ and the value at all other indices is the same as in $A$. The evaluation of a simple program in our language is shown below:\\

$(\lambda A. \; \tget(A, 2) + \tget(A,3)) \tset(\ttab(\lambda i . \; i, 5), 2, 10)$

$(\lambda A. \; \tget(A, 2) + \tget(A,3)) \tset([1,2,3,4,5], 2, 10)$

$(\lambda A. \; \tget(A, 2) + \tget(A,3)) [1,10,3,4,5]$

$\tget([1,10,3,4,5], 2) + \tget([1,10,3,4,5],3) $

@@

The dynamics given in the previous section are functional. However, the costs of get and set depend on whether the array is a leaf or internal array. To deal with this, we thread a store through the cost dynamics to keep track of whether each array is a leaf or internal array.

@@

$$\frac{\begin{gathered}\delta; e_1 \Downarrow \delta'; \lambda x . e; w_1 ; s_1 \;\;\; \delta'; e_2 \Downarrow \delta'';v; w_2 ; s_2 \;\;\; \\ \delta''; [v/x]e \Downarrow \delta'''; v'; w_3 ; s_3\end{gathered}}{\delta; e_1 \; e_2 \Downarrow \delta'''; v'; 1+w_1+w_2+w_3; 1+s_1+s_2+s_3} \text{ (func-app)}$$

@@

Arrays are represented by $(l, V)$ where $l$ is a label that uniquely identifies the array, and $V$ contains the values in the array. $V[i]$ is the $i^{\text{th}}$ value in the array. If \set{} has been called on an array, the array is an interior array (represented by $-$), otherwise the array is a leaf array (represented by $+$).

Array functions have different costs depending on whether the array is a leaf or interior array. We thread a store $\delta$ through each operation. $\delta$ maps array labels $l$ to $(+/-, c)$, where $c$ represents the number of cheap \get{}s on the array. $L(\delta)$ denotes the set of labels in the store $\delta$. Let $f(n)$ be the work of calling \get{} on an interior array of size $n$ and $g(n)$ be the work of calling \set{} on an interior array of size $n$.

The cost dynamics for the functions on a functional array of size $n$ are:
$$\frac{A = (l,V) \;\;\; i \text{ val} \;\;\; v \text{ val} \;\;\; \delta[l \mapsto (+, c)] \;\;\; l' \not\in L(\delta)}{\delta; \tset \; A \; i \; v \Downarrow \delta[l \mapsto (-, c), l' \mapsto (+, 0)]; (l', V[i \mapsto v]); 1; 1} \text{  (set-leaf)}$$

$$\frac{A = (l,V) \;\;\; i \text{ val} \;\;\; v \text{ val} \;\;\; \delta[l \mapsto (-, c)] \;\;\;  l' \not\in L(\delta)}{\delta; \tset \; A \; i \; v \Downarrow \delta[l' \mapsto (+, 0)]; (l', V[i \mapsto v]); g(n); 1} \text{  (set-interior)}$$

$$\frac{\begin{gathered}\delta; e_1 \Downarrow \delta_1; A; w_1; s_1 \;\;\; \delta_1; e_2 \Downarrow \delta_2; i; w_2; s_2 \;\;\; \\ \delta_2; e_3 \Downarrow \delta_3; v; w_3; s_3 \;\;\; \delta_3; \tset \; A \; i \; v \Downarrow \delta'; A'; w'; s' \end{gathered}}{\delta; \tset \; e_1 \; e_2 \; e_3 \Downarrow \delta'; A'; w_1+w_2+w_3+w'; s_1+s_2+s_3+s'} \text{ (set-eval)}$$

$$\frac{A = (l,V) \;\;\; i \text{ val} \;\;\; \delta[l \mapsto (+, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta[l \mapsto (+, c+1)]; V[i]; 1; 1} \text{  (get-leaf)}$$

$$\frac{A = (l,V) \;\;\; i \text{ val} \;\;\; \delta[l \mapsto (-, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta; V[i]; f(n); 1} \text{  (get-interior)}$$

$$\frac{\begin{gathered}\delta; e_1 \Downarrow \delta_1; A; w_1; s_1 \;\;\; \delta_1; e_2 \Downarrow \delta_2; i; w_2; s_2 \;\;\; \\ \delta_2; \tget \; A \; i \Downarrow \delta'; v'; w'; s' \end{gathered}}{\delta; \tget \; A \; i \Downarrow \delta'; v'; w_1 + w_2 + w'; s_1 + s_2 + s'}\text{ (get-eval)}$$

$$\frac{\begin{gathered}\delta, e_L \Downarrow \delta_L; v_L; w_L, s_L \;\;\; \delta, e_R \Downarrow \delta_R; v_R; w_R, s_R \;\;\; \\ (L(\delta_L) \setminus L(\delta)) \cap (L(\delta_R) \setminus L(\delta)) = \emptyset\end{gathered}}{\delta; (e_L || e_R) \Downarrow \delta'; (v_L || v_R); 1 + w_L + w_R + w'; \max(s_L, s_R)} \text{  (fork-join)}$$

\set{} creates a new label and array and modifies the store to indicate that the newly created array is a leaf array, and the array \set{} was called on has become an interior array. The work for \set{} is higher for interior arrays. \get{} returns the value at the desired index, if the array is a leaf then the operation costs 1 unit of work but we increment the counter of cheap \get{}s in the store.

@@

Further, suppose that $(\delta_L \setminus \delta) \cap (\delta_R \setminus \delta) = \emptyset$ (the new labels produced on both sides of the fork-join do not conflict). Consider an array $A = (l, V)$ with $\delta[l] = (s,c)$, $\delta_L[l] = (s_L, c + c_L)$, $\delta_R[l] = (s_R, c + c_R)$. When multiplying two signs or multiplying a sign with an integer, consider + to be 1 and - to be 0.

\func{Combine} describes how to combine the values in the store on both sides of the fork-join. The array is a leaf array iff it is a leaf array on both sides of the fork-join. Cheap \get{}s on one side of the fork remain cheap iff there were no calls to \set{} on the other side of the fork.
\begin{equation*}
  \begin{aligned}
   \text{\func{combine}}&((s, c), (s_L, c + c_L), (s_R, c + c_R))= \\
   &(s_L s_R, c + s_R c_L + s_L c_R)
  \end{aligned}
\end{equation*}
$$\delta' = \; (\delta_L / \delta) \; \cup \; (\delta_R / \delta) \; \cup \bigcup_{l \in L(\delta)} [l \mapsto \text{\func{combine}}(\delta[l], \delta_L[l], \delta_R[l])]$$
\func{Extrawork} gives the additional cost incurred if an array of size $n$ was modified on either side of a fork-join. If the array was an interior array before the fork-join, then all functions incurred their maximal cost and there is no additional cost. Otherwise, cheap \get{}s on one side of the fork are charged $f(n)$ work iff the other side of the fork called \set{}. If both sides of the fork called \set{}, then one of the \set{}s came first and has work $1$, the subsequent \set{} has work $g(n)$.
\begin{equation*}
  \begin{aligned}
  \text{\func{extra}}&\text{\func{work}}(l, (s, c), (s_L, c + c_L), (s_R, c + c_R)) = \\
  & ((\neg s_R) c_L  + (\neg s_L) c_R)(f(size(l))-1) + \\
  &s(\neg s_R) (\neg s_L) (g(size(l))-1)
 \end{aligned}
\end{equation*}
$$w' = \sum_{l \in L(\delta), \delta[l \mapsto (+, c)]} \text{\func{extrawork}}(l, \delta[l], \delta_L[l], \delta_R[l])$$
Then, the fork-join cost semantics are:
$$\delta; (e_L || e_R) \Downarrow \delta'; (v_L || v_R); 1 + w_L + w_R + w'; \max(s_L, s_R)$$

\tabulate{} involves multiple function calls in parallel, so its cost semantics use \func{combine} and \func{extrawork} multiple times. We omit discussion of \tabulate{} here and in the rest of the paper because it involves messy details that are not relevant to the goals of this paper.

\section{Analysis}

\subsection{Interleaved Structural Dynamics}

To show that our implementation is correct and to validate our cost dynamics, we introduce interleaved structural dynamics for the source language. The structural dynamics describe the possible steps a program in our source language can take and are given by the following judgement, where $S$ is the store, $e$ is the expression, $S'$ is the resulting store, $e'$ is the resulting expression after a step has been taken, and $w$ is the work.

$$S, e \to S', e', w$$

For now we let the span of each step be 1 (we account for the spans later on).

As before, arrays are represented by $(l, V)$ where $l$ is a unique label and $V$ contains the values in the array. However, the store $S$ maps labels only to $+$ (leaf) or $-$ (interior). $L(S)$ denotes the set of labels in the store $S$. \func{size}$(l)$ denotes the size of the array referenced by label $l$. The judgement \func{val} describes terminal values. The structural dynamics are given as follows:
$$\frac{e_2 \; \text{val}}{S, (\lambda x.e)e_2 \to S, [e_2/x]e, 1} \text{ (app-sub)}$$
$$\frac{S, e_2 \to S', e_2', w}{S, (\lambda x.e) e_2 \to S', (\lambda x.e) e_2', w} \text{ (app-arg)}$$
$$\frac{S, e_1 \to S', e_1', w}{S, e_1 e_2 \to S', e_1' e_2, w} \text{ (app-func)}$$
$$\frac{}{S, \text{if } true \text{ then } e_2 \text{ else } e_3 \to S, e_2, 1} \text{ (if-true)}$$
$$\frac{}{S, \text{if } false \text{ then } e_2 \text{ else } e_3 \to S, e_3, 1}  \text{ (if-false)}$$
$$\frac{S, e_1 \to S', e_1', w}{S, \text{if } e_1 \text{ then } e_2 \text{ else } e_3 \to S', \text{if } e_1' \text{ then } e_2 \text{ else } e_3, w} \text{ (if-cond)}$$
$$\frac{A = (l, V) \;\;\; S[l] = + \;\;\; i \text{ val}}{S, get(A, i) \to S, V[i], 1} \text{ (get-leaf)}$$
$$\frac{A = (l, V) \;\;\; S[l] = -  \;\;\; i \text{ val}}{S, get(A, i) \to S, V[i], f(size(l))} \text{ (get-interior)}$$
$$\frac{A = (l, V) \;\;\; S, e_2 \to S', e_2', w}{S, get(A, e_2) \to S', get(A, e_2'), w} \text{ (get-index)}$$
$$\frac{S, e_1 \to S', e_1', w}{S, get(e_1, e_2) \to S', get(e_1', e_2), w} \text{ (get-array)}$$
$$\frac{A = (l, V) \;\;\; S[l] = + \;\;\;  l' \not\in L(S) \;\;\; i \text{ val} \;\;\; v \text{ val}}{S, set(A, i, v) \to S[l \mapsto -], (l', V[i \mapsto v]), 1} \text{ (set-leaf)}$$
$$\frac{A = (l, V) \;\;\; S[l] = - \;\;\;  l' \not\in L(S) \;\;\; i \text{ val} \;\;\; v \text{ val}}{S, set(A, i, v) \to S, (l', V[i \mapsto v]), g(size(l))} \text{ (set-interior)}$$
$$\frac{A = (l, V) \;\;\; i \text{ val} \;\;\; S, e_3 \to S', e_3', w}{S, set(A, i, e_3) \to S', set(A,i,e_3'), w} \text{ (set-val)}$$
$$\frac{A = (l, V) \;\;\; S, e_2 \to S', e_2', w}{S, set(A, e_2, e_3) \to S', set(A,e_2',e_3), w} \text{ (set-index)}$$
$$\frac{e_1 \to S', e_1', w}{S, set(e_1, e_2, e_3) \to S', set(e_1',e_2,e_3), w} \text{ (set-array)}$$

The structural dynamics for fork-join is non-deterministic since either side of the fork can take a step.
$$\frac{S, e_1 \to S', e_1', w}{S, e_1 || e_2 \to S', e_1' || e_2, w} \text{ (fork-left)}$$
$$\frac{S, e_2 \to S', e_2', w}{S, e_1 || e_2 \to S', e_1 || e_2', w} \text{ (fork-right)}$$

The structural dynamics for tabulate is also non-deterministic.

\begin{definition}
A \emph{transition sequence} $T$ is a sequence of states $[(S_0, e_0), ..., (S_n, e_n)]$ and costs $[w_1, ..., w_n]$ s.t. for all $0 \leq i < n$, $S_i, e_i \to S_{i+1}, e_{i+1}, w_{i+1}$ . We say that $T$ takes $S_0, e_0$ to $S_n, e_n$ and has length $n$ and denote this by $S_0, e_0 \to^n S_n, e_n$.
\end{definition}

\begin{definition}
We say that $T$ is \emph{maximal} if there does not exist $S', e', w'$ s.t. $S_n, e_n \to S', e', w'$.
\end{definition}

\begin{definition}
A \emph{transition subsequence} $T_{i, j}$ of $T$ is given by the sequence of states $[(S_i, e_i), ..., (S_j, e_j)]$ and costs $[w_{i+1}, ..., w_j]$. Note that $T = T_{0,n}$.
\end{definition}

\begin{definition}
The \emph{work of a transition sequence} $T$ is given by:
$$W(T) = \sum_{i=1}^n w_i$$
Different transition sequences starting and ending at the same states may have different work.
\end{definition}

\subsection{Sequential Cost Analysis}

\begin{definition}
Consider a transition sequence $T$. We say that $S_i, e_i \to S_{i+1}, e_{i+1}, w_{i+1}$ is a get on $l$ if the step involves evaluating the get function on array $(l, V)$. We say it is a \emph{cheap get} on $l$ if $w_{i+1} = 1$. The number of cheap gets on $l$ in $T$ is denoted by $SC_l(T)$.
\end{definition}

\begin{definition}
Unlike the store in the evaluational dynamics, the store in the structural dynamics only stores whether each array is $+$ (leaf) or $-$ (interior). $sign(\delta)$ takes a store that maps $l \mapsto (s,c)$ and returns a store which maps $l \mapsto s$.
\end{definition}

\begin{definition}
Suppose that $\delta, e \Downarrow \delta', e', w', s'$. We define the number of cheap gets in going from $\delta$ to $\delta'$ in the evaluational dynamics as follows. Suppose $l \mapsto (s', c') \in \delta'$. If $l \mapsto (s,c) \in \delta$ then $EC_l(\delta, \delta') = c' - c$ and if $l \not\in \delta$ then $EC_l(\delta, \delta') = c'$. If $l \not\in \delta'$ then $EC_l(\delta, \delta') = 0$.
\end{definition}

\begin{definition}
A \emph{relabeling} of $L_1$ is a bijective function $R$ from label sets $L_1 \to L_2$. $R$ can be used to relabel stores and expressions. $R(\delta) = \{ R(l) \mapsto e \; | \; l \mapsto e \in \delta \wedge l \in L_1  \} \cup \{ l \mapsto e \; | \; l \mapsto e \in \delta \wedge l \not\in L_1  \}$. In other words, $R$ relabels some of the labels in $\delta$. Similarly, $R(e)$ returns an expression $e'$ where each occurrence of a label $l$ in $e$ is substituted by $R(l)$.
\end{definition}

\begin{lemma}
\label{relabeling_lemma}
Suppose that there exists a derivation of length $m$ that $\delta, e \Downarrow \delta', e', w, s$. Let $R$ be an arbitrary labeling. Then there exists a derivation of length $m$ that $R(\delta), R(e) \Downarrow R(\delta'), R(e'), w, s$.
\end{lemma}

\begin{proof}
By applying the relabeling to each step of the derivation, and noting that all rules in the evaluational dynamics hold under relabelings.
\end{proof}

\begin{lemma}
\label{stateless_lemma}
If $S, e \to S', e', w$ and $U$ is a store s.t. for all $l \in e$, $l \in U$, then there exists $U', w'$ s.t. $U, e \to U', e', w'$
\end{lemma}

\begin{proof}
By casing on each rule in the structural dynamics, and noting that transitions of expressions never depend on the values in the store.
\end{proof}

\begin{theorem}
\label{main_work_theorem}
Suppose that 
$$\delta, e_0 \Downarrow \delta_E, e_E, w_E, s_E$$

and consider arbitrary maximal transition sequence T taking 
$$S_0 = sign(\delta), e_0 \to^n S_n, e_n$$
å
The following hold:
\begin{enumerate}
\item For some relabeling $R$ of $L(\delta_E) \setminus L(\delta)$, $S_n = sign(R(\delta_E))$ and $e_n = R(e_E)$.
\item For all $l \in \delta_E$, $EC_l(\delta, \delta_E) \leq SC_{R(l)}(T)$.
\item The sum of work and cheap gets is conserved: \begin{align*}
w_E + &\sum_{l \in L(\delta_E)} EC_l(\delta, \delta_E) (f(size(l))-1) =\\
 &W(T) + \sum_{l \in L(\delta_E)} SC_{R(l)}(T) (f(size(R(l))) - 1)
 \end{align*}
\end{enumerate}
\end{theorem}

\begin{proof}
By induction on the length of the shortest derivation in the evaluation dynamics. We prove the theorem for 2 of the rules in the evaluation dynamics: get-leaf and fork-join. The other cases follow from a similar line of reasoning.
\\

\noindent \textbf{Case get-leaf}: Suppose $e_0 = get(e_L, e_R)$ and
$$\delta, e_L \Downarrow \delta_L, e_L', w_L$$
$$\delta_L, e_R \Downarrow \delta_R, e_R', w_R$$
$$\delta_R, get(e_L', e_R') \Downarrow \delta_E, e_E, 1$$
which gives us
$$\delta, get(e_L, e_R) \Downarrow \delta_E, e_E, w_E$$
with $w_E = w_L + w_R + 1$.

Note that the structural dynamics first step the left argument to get, then the right argument, and finally evaluate the get. So there exists $m$ such that $T_{0,m}$ involves stepping the left argument, $T_{m,n-1}$ involves stepping the right argument, and $T_{n-1,n}$ involves evaluating the get.

\textbf{Part 1}: Suppose that in $T_{0,m}$, for $0 \leq i \leq m$, $e_i = get(e_{L,i}, e_R)$. Consider a new projected transition sequence $T_{0,m}'$ with states $[(S_0, e_{L,0}), ..., (S_m, e_{L,m})]$ and costs the same as $T_{0,m}$: $[w_1, ..., w_m]$. It is easy to verify that $T_{0,m}'$ is a valid transition sequence starting at $(S_0, e_{L,0})$ and is maximal.

$S_0 = sign(\delta)$ and $e_{L,0} = e_L$ so we can apply the IH to $T_{0,m}'$. For some relabeling $R$ of labels in $L(\delta_L) / \delta$, $s_m = sign(R(\delta_L))$ and $e_{L,m} = R(e_L')$ $\Rightarrow$ $e_m' = (e_{L,m}, e_R) = (R(e_L'), e_R)$. WLOG suppose that $\delta_L$ was labeled such that the above hold (we can do this because of lemma \ref{relabeling_lemma}). Note that in subsequent parts of the proof, we omit creating the projected transition sequence and apply the IH directly to $T_{0,m}$ when needed.

Using similar logic on $T_{m,n-1}$ we get that for some relabeling of $\delta_R / \delta_L$, applying that relabeling to $\delta_R$ and $(e_L', e_R')$ gives us $e_{n-1} = (e_L', e_R')$ and $S_{n-1} = sign(\delta_R)$. WLOG suppose that $\delta_R$ was labeled so that the above holds.

By combining the relabelings of $\delta_L / \delta$ and $\delta_R / \delta_L$ and applying that to $(e_L', e_R')$ and $\delta_R$, we get that $e_{n-1} = get(e_L', e_R')$ and $S_{n-1} = sign(\delta_R)$. Note that get does not change the sign of any label in the store in either the structural or the evaluation dynamics, so $sign(\delta_E) = sign(\delta_R) = S_{n-1} = S_n$. Further, the structural and evaluational dynamics apply get in the same way, so $e_n = e_E$. Therefore the first part of this theorem holds. WLOG suppose that the stores and expressions were relabeled s.t. the first part holds.

\textbf{Part 2}: From lemma \ref{relabeling_lemma}, the length of the shortest derivation for the relabeled evaluational dynamics does not change, so we can apply the inductive hypothesis. Applying the IH to $T_{0,m}$, for all $l \in \delta_L$, $EC_l(\delta, \delta_L) \leq SC_l(T_{0,m}') = SC_l(T_{0,m})$. If $l \in \delta_R$ but $l \not\in \delta_L$ then $EC_l(\delta, \delta_L) = SC_l(T_{0,m}) = 0$ (intuitively the label did not exist and so did not have any cheap gets), so in particular, for all $l \in \delta_R$, $EC_l(\delta, \delta_L) \leq SC_l(T_{0,m})$. Applying the IH to $T_{m,n-1}$ we get that for all $l \in \delta_R$, $EC_l(\delta_L, \delta_R) \leq SC(T_{m,n-1})$.

We note that $EC$ and $SC$ are additive, that is,
$$EC_l(\delta, \delta_R) = EC_l(\delta, \delta_L) + EC_l(\delta_L, \delta_R)$$
$$SC_l(T_{0,n-1}) = SC_l(T_{0,m}) + SC_l(T_{m,n-1})$$
This implies that for all $l \in \delta_R$, $EC_l(\delta, \delta_R) \leq SC_l(T_{0,n-1})$.

Now, let $e_L' = (l',V)$ with $l' \in \delta_R$. Then, $EC_{l'}(\delta, \delta_E) = EC_{l'}(\delta, \delta_R) + 1$ and $SC_{l'}(T) = SC_{l'}(T_{0,n-1}) + 1$, so $EC_{l'}(\delta, \delta_E) \leq SC_{l'}(T)$. If $l \neq l'$ then $EC_{l}(\delta, \delta_E) = EC_{l}(\delta, \delta_R)$ and $SC_{l}(T) = SC_{l}(T_{0,n-1})$, so $EC_{l}(\delta, \delta_E) \leq SC_{l}(T)$. So the second part of the theorem holds.

\textbf{Part 3}: The third part of the theorem follows by similarly applying the IH on $T_{0,m}$ and $T_{m,n-1}$, and noting that $W(T) = W(T_{0,m}) + W(T_{m,n-1}) + 1$.
\\

\noindent \textbf{Case fork-join}: Suppose $e_0 = (e_L || e_R)$ and
$$\delta, e_L \Downarrow \delta_L, v_L, w_L$$
$$\delta, e_R \Downarrow \delta_R, v_R, w_R$$
which gives us
$$\delta, (e_L || e_R) \Downarrow \delta', (v_1 || v_2), 1 + w_L + w_R + w'$$
with $w'$ and $\delta'$ defined in terms of \func{extrawork} and \func{combine} as given in the evaluational dynamics.

In T, for all $0 \leq i \leq n$ let $e_i = (e^L_i || e^R_i)$. Let $I_L$ be the sequence of all $0 \leq i \leq n-1$ where transitioning from $e_i$ to $e_{i+1}$ involved stepping the left argument. Let $m$ be the length of $I_L$. Similarly define $I_R$ for the right argument, and let $k$ be the length of $I_R$.
